\plaintitle

### 2016-09-21 Paper/figures/figures.tex
Dumping more figures, initial stab at implications for design.

>>>>

\newcommand{\expThreeErrorFig}{
	\begin{figure}
		\centering
		\includegraphics[width=.85\columnwidth]{figures/exp3errorType}
		\caption{From experiment 3, the absolute error in estimations of trend defined in two ways: either as difference from the robust slope (which excludes outliers), or the OLS slope (which does not). Participants were significantly closer to the robust fit, indicating that they were largely insensitive to outliers. We include the error from trials with no outliers (and so where the two methods of calculating error agree) for reference. Error to the robust line is higher than this standard, indicating that participants were at least performing some interpolation between OLS and robust fits, even if they largely favored the robust fit. Confidence intervals represent bootstrapped 95\% c.i.s of interquartile mean. }
		\label{fig:exp3errorType}
	\end{figure}
}

### 2016-09-21 Paper/proceedings.tex
Dumping more figures, initial stab at implications for design.

In this work, we describe a series of crowdsourced experiments on the visual estimation of trends in common bivariate visualizations such as scatter plots, area charts, and line graphs. We present the results of three studies investigating estimation of trend slope, trend intercept, and the effect of outliers. We find that, while in most cases viewers accurately estimate trends, area charts introduce systematic underestimation of trend intercept, and that viewers give low weights to extreme values when estimating trends. These results suggest that there are several areas where human judgments diverge from the fitted models generated by techniques such as Ordinary Least Squares (OLS), and that the design of bivariate visualizations can introduce additional biases in these judgments. Designers should therefore

DO SOMETHING

@@

\outliersFig
OLS regression operates under the assumption that there is a unimodal, symmetric distribution of residuals surrounding the line of best fit. Extreme outliers violate this assumption, and can result in trend lines that are substantially different from those produced by more robust methods. Visual inspection can accurately identify certain classes of outliers \cite{albers2014task}. Therefore, regression by eye may afford the estimation of both outlier-robust and outlier-sensitive trends. However, cognitive biases~\cite{tversky1975judgment} such as anchoring (the tendency to overweight the first set of information), availability (the tendency to overweight more recent or extreme information), and the hot-hand fallacy (the tendency to assume that trends of high or low values with continue) can impact how viewers categorize and utilize outliers in prediction tasks ~\cite{campbell2009anchoring, ji2001culture}. That is, regression by eye may not uniformly weight outliers, depending on their position in the plot.

@@

@@

Our results support our first hypothesis. \textbf{Participant estimates were closer to the robust trend line than the trend line that included outliers}. Excluding conditions with zero outliers (and so the two trend lines would be identical), the interquartile mean of absolute error was over $4$ times higher when calculated as a comparison to the OLS trend line ($0.36$) as opposed to the robust trend line ($0.08$). A Student's T-test confirmed that the error as defined by the OLS line was significantly higher than the robust line ($t(6670)= 43$, $p<0.001$). However, a Student's T-test found a significant difference in unsigned error when compared to the robust trend line between estimates where no outliers were present, and those with any amount of outliers ($\mu=0.07$ vs. $\mu=0.08$, $t(3808) = 9.8$, $p<0.001$). This indicates that participants are not entirely ignoring outliers, although the small effect size suggests that participants are giving outliers significantly less weight than OLS regression. Figure \ref{fig:exp3outliers} illustrates this result.

Our results fail to support our second hypothesis. \textbf{While participant estimates diverge from the robust line as the number of outliers increases, the OLS trend line diverges faster than participant estimates}. There was a significant positive interaction between outlier count and absolute error with respect to the robust trend line ($R^2 = 0.02$,$F(1,4591)=110$, $p<0.001$). However, there was also a significant positive interaction between outlier count and absolute error with respect to the OLS line  ($R^2 = 0.38$,$F(1,4591)=2800$, $p<0.001$). Additionally, while the range of errors with respect to the robust line remained small (interquartile means of $0.07$ with 0 outliers, monotonically increasing to $0.10$ with 15 outliers), absolute error with respect to the OLS line increased almost 7-fold, from $0.07$ with 0 outliers, monotonically to $0.54$ with 15 outliers. These results indicate that while increased numbers of extreme values may result in participants increasing the weight of these outliers, this increase in weight is slower than the sensitivity of OLS to outliers.

>>>>

In this work, we describe a series of crowdsourced experiments on the visual estimation of trends in common bivariate visualizations such as scatter plots, area charts, and line graphs. We present the results of three studies investigating estimation of trend slope, trend intercept, and the effect of outliers. We find that, while in most cases viewers accurately estimate trends, area charts introduce systematic underestimation of trend intercept, and that viewers give low weights to extreme values when estimating trends. These results suggest that there are several areas where human judgments diverge from the fitted models generated by techniques such as Ordinary Least Squares (OLS), and that the design of bivariate visualizations can introduce additional biases in these judgments. Designers should therefore make an informed decision between designing for regression by eye and the explicit annotation of statistical regression information. Regression by eye affords quick estimations and flexibility in model selections, but is subject to perceptual biases and inaccuracies in estimation. Explicit annotation affords accuracy, but constrains the model space, and adds visual complexity to plots.

@@

\outliersFig

OLS regression operates under the assumption that there is a unimodal, symmetric distribution of residuals surrounding the line of best fit. Extreme outliers violate this assumption, and can result in trend lines that are substantially different from those produced by more robust methods. Visual inspection can accurately identify certain classes of outliers \cite{albers2014task}. Therefore, regression by eye may afford the estimation of both outlier-robust and outlier-sensitive trends. However, cognitive biases~\cite{tversky1975judgment} such as anchoring (the tendency to overweight the first set of information), availability (the tendency to overweight more recent or extreme information), and the hot-hand fallacy (the tendency to assume that trends of high or low values with continue) can impact how viewers categorize and utilize outliers in prediction tasks ~\cite{campbell2009anchoring, ji2001culture}. That is, regression by eye may not uniformly weight outliers, depending on their position in the plot.

@@

\expThreeErrorFig

@@

Our results support our first hypothesis. \textbf{Participant estimates were closer to the robust trend line than the trend line that included outliers}. Excluding conditions with zero outliers (and so the two trend lines would be identical), the interquartile mean of absolute error was over $4$ times higher when calculated as a comparison to the OLS trend line ($0.36$) as opposed to the robust trend line ($0.08$). A Student's T-test confirmed that the error as defined by the OLS line was significantly higher than the robust line ($t(6670)= 43$, $p<0.001$). However, a Student's T-test found a significant difference in unsigned error when compared to the robust trend line between estimates where no outliers were present, and those with any amount of outliers ($\mu=0.07$ vs. $\mu=0.08$, $t(3808) = 9.8$, $p<0.001$). Figure \ref{fig:exp3errorType} illustrates this result. This indicates that participants are not entirely ignoring outliers, although the small effect size suggests that participants are giving outliers significantly less weight than OLS regression. Figure \ref{fig:outlier} shows an example of this weighting.

Our results fail to support our second hypothesis. \textbf{While participant estimates diverge from the robust line as the number of outliers increases, the OLS trend line diverges faster than participant estimates}. There was a significant positive interaction between outlier count and absolute error with respect to the robust trend line ($R^2 = 0.02$,$F(1,4591)=110$, $p<0.001$). However, there was also a significant positive interaction between outlier count and absolute error with respect to the OLS line  ($R^2 = 0.38$,$F(1,4591)=2800$, $p<0.001$). Additionally, while the range of errors with respect to the robust line remained small (interquartile means of $0.07$ with 0 outliers, monotonically increasing to $0.10$ with 15 outliers), absolute error with respect to the OLS line increased almost 7-fold, from $0.07$ with 0 outliers, monotonically to $0.54$ with 15 outliers. These results indicate that while increased numbers of extreme values may result in participants increasing the weight of these outliers, this increase in weight is slower than the sensitivity of OLS to outliers.  Figure \ref{fig:exp3outliers} illustrates this result.
